
\chapter{Junicode on the Web}\hypertarget{OnTheWeb}{}
%\fancyhead[CE]{\scshape\color{myRed} {\addfontfeatures{Numbers=OldStyle}\thepage}\hspace{10pt}on the web}

If you are using Junicode on a web page, you should prefer the variable fonts
(those with “VF” in the family name and filename) to the static fonts. One
variable font file can do the work of many traditional font files. For example,
the \href{https://psb1558.github.io/Junicode-font/Junicode-2-feature-test.html}%
{Test of High-Level CSS Properties} web page displays Junicode in regular,
bold and semicondensed styles. It used
to be that your user would have to download three font files, one for each
style, but one variable font will display all three.

But you may be thinking, \textit{That font is big}! It's true: even the
compressed webfont (.woff2) is nearly a megabyte in size—enough to seriously
slow down the loading of a web page.

To solve this problem, you'll need to subset the font—that is, produce a copy
that contains only what you need. The subsetted font that downloads
with the property test web page
is approximately 275k in size—almost thirty percent of the
size of the full webfont. It's still a pretty big download, but that's because the page
displays a lot of the font's features. If I were displaying, say, a diplomatic
transcript of a Latin text, the font would be much smaller.

So the first section of this chapter will talk about how to subset the Junicode
font.

\section{Subsetting Junicode}

First the legalities. It is perfectly all right to create a modified version
of Junicode via subsetting, compress it into a webfont (almost certainly in
woff2 format), and host it on your web server. This is because “Junicode” is
not a “Reserved Font Name” (which complicates web use of
many fonts licensed under the Open Font License). If you are nervous about the
legal requirements of the Open Font License, you can change the font name to
something arbitrary with the \texttt{-‌-obfuscate-names} option
of the pyftsubset program, and you can embed the Open Font License, or a
link to it, in your CSS. These steps should settle any ambiguity about whether
you are in compliance with the license.

Generating a subsetted version of Junicode should be one of the last tasks you
perform before deploying your web page(s). Until then, it is recommended that
you work with the unmodified font. After subsetting, review your
pages thoroughly to make sure everything is displayed properly. If you have
forgotten to include a glyph in your subsetted font, you will see little boxes
where characters should be or, perhaps, the correct characters
displayed in the wrong typeface. If you have omitted features, you will see
default instead of transformed characters.

There are many subsetting programs, some online and very easy to use. But for
maximum control (and thus the smallest fonts), you should choose pyftsubset, a
part of the \href{https://github.com/fonttools/fonttools}{fontTools} library,
which runs under Python 3.7 or higher. This is a command-line tool which
takes a long list of arguments; you should create a shell script to run it.

Here is the script used to create the subsetted font for the property test
web page mentioned above:

\small\begin{verbatim}
#!/bin/zsh
pyftsubset JunicodeVF-Roman.ttf \
--flavor=woff2 --output-file=JunicodeVFsubset.woff2 \
--recommended-glyphs \
--text="jq" --text-file=Junicode-2-feature-test.html \
--layout-features+=liga,ss01,ss02,ss03,ss04,ss05,ss06,ss07,ss08,\
ss10,ss12,ss13,ss14,ss15,ss16,ss17,ss18,ss19,ss20,cv01,cv02,cv05,\
cv06,cv07,cv08,cv09,cv10,dlig,hlig,onum,pnum,pcap,smcp,c2sc,subs,\
sups,zero \
--layout-features-=rlig
\end{verbatim}

\normalsize\noindent For those unfamiliar with shell scripts, the first line specifies the shell
the script is to run under (in this case the default shell for Mac OS, but
\texttt{bash} is another possible choice), and the backslashes mean
that the command continues on the next line. The rest of the file is a list
of arguments passed to pyftsubset. Let's walk through them.

First after the program name comes the name of the unsubsetted, uncompressed
font file. After that,
the \texttt{--flavor} argument tells the program that you want a webfont in
woff2 format, and \texttt{--output} is the name of the font file you want the
program to save.

Having taken care of this preliminary business, we tell pyftsubset what we
want the font to contain.

\texttt{--recommended-glyphs} includes a few characters that
every font should have, according to the OpenType specification---though in
fact modern browsers don't care. It's best, however, to conform to the specification,
since it's impossible to say with absolute certainty that no program will
ever reject the font
because of the absence of these few characters.

\texttt{--text-file} is the name of a file to treat as a list of characters
that \textit{must} be included in the font. In this case I have simply used
the html file for the web page for this purpose. If your site contains
multiple web pages, your job will be more complicated. You must make sure the
text file contains all the characters used on the site---either that or
supplement the text file with a \texttt{--text} argument (which here adds
two lowercase letters that don't appear in the web page---just in case).
The text file will
contain only encoded characters—you don't have to worry here about unencoded
characters produced by OpenType features.

\texttt{--layout-features+} tells the program which OpenType features you
want to retain in the font. All others, except for the
\hyperlink{req}{Required Features}, are discarded. All of the characters
referenced in these features will also be included in the output file, as long
as those characters are variants of characters in your text file. For example,
the \textSourceText{smcp} (Small Caps) feature has many more small caps than there are letters
of the alphabet, but most of them are not included in the subsetted font. The
program's parsimony with characters keeps the font file as small as possible.
Note that some features are included automatically: \textSourceText{ccmp},
\textSourceText{locl}, \textSourceText{calt}, \textSourceText{liga}, \textSourceText{rlig},
\textSourceText{kern}, \textSourceText{mark}, and \textSourceText{mkmk}.


\texttt{--layout-features-} tells the program which OpenType features to omit.
Normally, \textSourceText{rlig} (Required Ligtures) is automatically
included in fonts by pyftsubset, but as it has no relevance to this web page,
it can be omitted.

These are the most useful arguments, but there are many more. Type
\texttt{pyftsubset --help} for a complete list. Once you have written your
script, run it (in Mac OS or Linux you need to make the file executable
by typing \texttt{chmod +x mysubsetscript} on the command line).

Before you put your subsetted font to work, check it carefully in a program
like \href{https://github.com/justvanrossum/fontgoggles}{FontGoggles},
which lets you preview the font and test all its OpenType features. If you
find errors, revise your script and run it again.

\section{Junicode and CSS/HTML}

This section assumes a basic knowledge of HTML (Hypertext Markup Language, used
to construct web pages) and CSS (Cascading Style Sheets, used to format them).
If you want to learn about these subjects, the number of good books and online
tutorials is so great that it makes no sense to try to list them. Just make sure
that the instructional materials you choose are of recent vintage, because the
relevant standards are always changing.

In the CSS for your web page, the \texttt{@font-face} at-rule for a variable
font is a little different
from the one for a static font in that the range of possible values for each
axis can be declared:

\small\begin{verbatim}
@font-face {
    font-family: "Junicode VF";
    src: url("./webfiles/JunicodeVFsubset.woff2");
    font-weight: 300 700;
    font-stretch: 75% 125%;
    font-style: normal;
}
\end{verbatim}

\normalsize\noindent These ranges are not strictly necessary, but they will prevent your
supplying invalid values for \texttt{font-weight} and \texttt{font-stretch}
(that is, \texttt{width}) in other CSS rules.

Once you have declared the font, you can invoke it in setting up classes.
For example:

\small\begin{verbatim}
body {
  font-family: "Junicode VF";
  font-size: 28px;
  font-weight: normal; /* that is, 400 */
  font-stretch: 112.5%; /* that is, semiexpanded */
}
h1 {
  font-family: "Junicode VF";
  font-size: 125%;
  font-weight: 600; /* that is, semibold */
  font-stretch: 112.5%; /* that is, semiexpanded */
}
.annotation {
  font-size: 90%;
  font-weight: 300; /* that is, light */
  font-stretch: 87.5%; /* that is, semicondensed */
}
\end{verbatim}

\normalsize\noindent These classes should be tested in all browsers. If any fail to
display text properly, you can use \texttt{font-variation-settings}
instead of the high-level \texttt{font-weight} and \texttt{font-stretch}:

\small\begin{verbatim}
body {
  font-family: "Junicode VF";
  font-size: 28px;
  font-variation-settings: "wght" 400, "wdth" 112.5;
}
h1 {
  font-family: "Junicode VF";
  font-size: 125%;
  font-variation-settings: "wght" 600, "wdth" 112.5;
}
.annotation {
  font-size: 90%;
  font-variation-settings: "wght" 300, "wdth" 87.5;
}
\end{verbatim}

\normalsize\noindent To accommodate older browsers, you should make a selection of
Junicode static fonts, subset them, and include them in your CSS. For example,
if you need normal and bold weights of Junicode roman, your
\texttt{@font-face} at-rule may look like this:

\small\begin{verbatim}
@font-face {
    font-family: "Junicode VF";
    src: url("./webfiles/JunicodeVFsubset.woff2");
    font-weight: 300 700;
    font-stretch: 75% 125%;
    font-style: normal;
}
@font-face {
    font-family: "Junicode";
    src: url("./webfiles/Junicode-Regular.woff2");
    font-weight: 400;
    font-style: normal;
}
@font-face {
    font-family: "Junicode";
    src: url("./webfiles/Junicode-Bold.woff2");
    font-weight: 700;
    font-style: normal;
}
\end{verbatim}

\normalsize\noindent Now use \texttt{@supports} in your CSS rules to determine which
font gets downloaded:

\small\begin{verbatim}
body {
  font-family: "Junicode", serif;
}
@supports (font-variation-settings: normal) {
  body {
    font-family: "Junicode VF", serif;
  }
}
b {
  font-weight: 700;
}
@supports (font-variation-settings: "wght" 700) {
  b {
    font-variation-settings: "wght" 700;
  }
}
\end{verbatim}

\normalsize\noindent The variable version of Junicode will be downloaded only if
the browser supports it, and the static version will be downloaded only if
needed.
